---
output:
  html_document: default
---

# Jackknife test of leverage 

```{r, echo=FALSE}
library(knitr)
opts_knit$set(root.dir = "..") 
```

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, message=FALSE}
library(tidyverse)
library(rstanarm)
library(RColorBrewer)
library(assertthat)
library(ggsidekick) # devtools::install_github("seananderson/ggsidekick")
```

Let's leave out each single data point from each model fit and collect all the predictions and plot them. 

```{r}
d_long <- readRDS("data/generated/d_long.rds")
labels <- readRDS("data/generated/labels.rds")
```


```{r}
N <- 100
newdata <- group_by(d_long, response) %>%
  do({
    data.frame(ph_scaled = predict(poly(.$ph, 2), 
      seq(min(.$ph), max(.$ph), length.out = N))[,1],
      ph = seq(min(.$ph), max(.$ph), length.out = N),
      ph_scaled2 = predict(poly(.$ph, 2), seq(min(.$ph), max(.$ph), 
        length.out = N))[,2])
  })
newdata <- bind_rows(data.frame(newdata, nutrients = 0), 
  data.frame(newdata, nutrients = 1))

jackknife <- function(dat) {
  fits_jk <- plyr::ddply(dat, "response", function(x) {
    
    print(unique(x$response))
    
    newdata_this_response <- 
      as.data.frame(filter(newdata, response == unique(x$response)))
    
    if (x$log_transform[1]) {
      f <- log(value) ~ ph_scaled + ph_scaled2 + nutrients
    } else
      f <- value ~ ph_scaled + ph_scaled2 + nutrients
    
    if (unique(x$response) == "ulva_grazing") # none when no nutrients
      f <- log(value) ~ ph_scaled + ph_scaled2
    
    if (x$family[1] == "gaussian") {
      fam <- gaussian(link = "identity")
      prior_aux <- student_t(df = 3, 0, 5, autoscale = FALSE)
    }
    if (x$family[1] == "count") {
      fam <- neg_binomial_2(link = "log")
      prior_aux <- student_t(df = 3, 0, 20, autoscale = FALSE)
    }
    
    if (unique(x$re) == FALSE) {
      m <- stan_glm(f, data = x, iter = 800, chains = 4, family = fam,
        prior = normal(0, 3, autoscale = FALSE),
        prior_intercept = normal(0, 10, autoscale = FALSE),
        prior_aux = prior_aux)
    } else {
      f <- as.formula(paste(Reduce(paste, deparse(f)), "+ (1 | tank)"))
      m <- stan_lmer(f, data = x, iter = 800, chains = 4,
        prior = normal(0, 3, autoscale = FALSE),
        prior_intercept = normal(0, 10, autoscale = FALSE),
        prior_aux = prior_aux, cores = 4)
    }
    
    b <- broom::tidy(m$stanfit, rhat = TRUE, ess = TRUE)
    assert_that(max(b$rhat) < 1.05)
    assert_that(min(b$ess) > 100)
    
    pp <- posterior_linpred(m, newdata = newdata_this_response, re.form = NA)
    med <- apply(pp, 2, median)
    data.frame(med, newdata_this_response)
  })
}

d_long <- mutate(d_long, tank = as.numeric(as.factor(paste(ph, nutrients))))

jk <- plyr::ldply(unique(d_long$tank), 
  function(i) {
    data.frame(jackknife(dplyr::filter(d_long, tank != i)), tank_removed = i)})

saveRDS(jk, file = "data/generated/jk.rds")
```

And plot the jackknifed output. 

```{r}
jk <- readRDS("data/generated/jk.rds")
nojk <- data.frame(jackknife(d_long), tank_removed = NA)
jk2 <- inner_join(jk, unique(select(d_long, response, log_transform, family)))
jk2 <- jk2 %>% group_by(response) %>%
  mutate(med = ifelse(log_transform | family == "count", exp(med), med)) %>%
  left_join(labels) %>% 
  mutate(label = sub("\\([a-zA-Z0-9 /%]*\\)", "", label)) %>% 
  mutate(label = sub(" $", "", label)) %>% 
  mutate(label = sub("\\\\n", "", label)) %>% 
  mutate(label_ordered = paste(figure_panel, label, sep = " - "))

nojk2 <- inner_join(nojk, unique(select(d_long, response, log_transform, family)))
nojk2 <- nojk2 %>% group_by(response) %>%
  mutate(med = ifelse(log_transform | family == "count", exp(med), med)) %>%
  left_join(labels) %>% 
  mutate(label = sub("\\([a-zA-Z0-9 /%]*\\)", "", label)) %>% 
  mutate(label = sub(" $", "", label)) %>% 
  mutate(label = sub("\\\\n", "", label)) %>% 
  mutate(label_ordered = paste(figure_panel, label, sep = " - "))

cols <- c(brewer.pal(4, "Greys")[[3]], brewer.pal(4, "Blues")[[3]])
filter(jk2) %>% 
  ggplot(aes(ph, med, color = as.factor(nutrients), 
    group = as.factor(paste(tank_removed, nutrients)))) +
  geom_line(alpha = 0.45) +
  facet_wrap(~label_ordered, scales = "free_y") +
  scale_x_reverse() + theme_sleek() +
  scale_colour_manual(values = cols) +
  labs(colour = "Nutrients", fill = "Nutrients", y = "Response value", 
    x = expression(Acidification~(mean~seawater~pH[T]))) +
  guides(colour = guide_legend(override.aes = list(alpha = 1, lwd = 1))) +
  theme(strip.text.x = element_text(size = 9)) +
  geom_line(data = nojk2, lwd = 1.5, alpha = 1)

ggsave("figs/jackknife.pdf", height = 7.5, width = 10)
```

# Interactions 

Let's verify quantitatively that there is little evidence for an interaction with the linear effect of pH with nutrients for the response variables with fairly high probability of nutrient main effects. We don't see much visual evidence of these interactions. 

```{r}
fits_interactions <- plyr::dlply(d_long, "response", function(x) {
  print(unique(x$response))
  if (x$log_transform[1]) {
    f <- log(value) ~ ph_scaled + ph_scaled2 + nutrients + ph_scaled:nutrients
  } else
    f <- value ~ ph_scaled + ph_scaled2 + nutrients + ph_scaled:nutrients
  
  if (unique(x$response) == "ulva_grazing") # none when no nutrients
    f <- log(value) ~ ph_scaled + ph_scaled2
  
  if (x$family[1] == "gaussian") {
    fam <- gaussian(link = "identity")
    prior_aux <- student_t(df = 3, 0, 5, autoscale = FALSE)
  }
  #  if (x$family[1] == "gamma")
  #    fam <- gaussian(link = "identity")
  # fam <- Gamma(link = "log")
  if (x$family[1] == "count") {
    fam <- neg_binomial_2(link = "log")
    prior_aux <- student_t(df = 3, 0, 20, autoscale = FALSE)
  }
  
  if (unique(x$re) == FALSE) {
    stan_glm(f, data = x, iter = 2000, chains = 4, family = fam,
      prior = normal(0, 3, autoscale = FALSE),
      prior_intercept = normal(0, 10, autoscale = FALSE),
      prior_aux = prior_aux)
  } else {
    f <- as.formula(paste(Reduce(paste, deparse(f)), "+ (1 | tank)"))
    stan_lmer(f, data = x, iter = 2000, chains = 4,
      prior = normal(0, 3, autoscale = FALSE),
      prior_intercept = normal(0, 10, autoscale = FALSE),
      prior_aux = prior_aux, cores = 4)
  }
})

fitsl <- map(fits_interactions, function(x) as.data.frame(as.matrix(x)))
fitsl$ulva_grazing$nutrients <- NA
fitsl$ulva_grazing$`ph_scaled:nutrients` <- NA

s <- plyr::ldply(fitsl, function(x) {
  as_data_frame(x[, names(x) == "ph_scaled:nutrients"])
}) %>%
  rename(response = .id)
```

Here I will calculate credible intervals to plot:

```{r}
cis <- gather(s, parameter, value, -response) %>% 
  group_by(response, parameter) %>% 
  summarise(l = quantile(value, 0.025, na.rm = TRUE),
    m = quantile(value, 0.5, na.rm = TRUE),
    u = quantile(value, 0.975, na.rm = TRUE)) %>% 
  ungroup() %>% 
  left_join(labels) %>% 
  mutate(label = sub("\\([a-zA-Z0-9 /%]*\\)", "", label)) %>% 
  mutate(label = sub(" $", "", label)) %>% 
  mutate(label = sub("\\\\n", "", label)) %>% 
  mutate(label_ordered = paste(toupper(figure_panel), label, sep = " - "))

ggplot(data = cis) +
  geom_point(aes(x = rev(label_ordered), y = m), col = "grey30") +
  geom_segment(aes(x = rev(label_ordered), y = l, xend = rev(label_ordered), yend = u)) +
  coord_flip() +
  geom_hline(yintercept = 0, lty = 2) +
  theme_sleek() +
  ylab("Linear interaction coefficient (pH and nutrients)") +
  xlab("")

ggsave("figs/interaction.pdf", height = 4, width = 6)
```

