How would the hypothesized shift in the quadratic effect appear in our
regression modeling results considering that we are not letting the quadratic
term itself interact with nutrient levels?

Let's simulate some data that approximately matches the shape in the hypothesis
figure.

```{r}
x <- seq(0, 10, length.out = 100)
y1 <- (x - 5) ^ 2 * (-5)
y2 <- (x - 3) * 16 + (x - 3) ^ 2 * (-8) - 8
plot(x, y1, type = "l", lty = 1)
lines(x, y2, lty = 2)
```

Let's combine that simulated data into a data frame that we can fit a model to.
We will also add a little bit of noise to the response data so we can fit
a model to it.

```{r}
set.seed(1)
noise <- rnorm(length(x), 0, 0.8)
x_centered <- x - mean(x)
dat <- data.frame(x = c(x_centered, x_centered), y = c(y1 + noise, y2 + noise),
  nutrients = c(rep(FALSE, length(y1)), rep(TRUE, length(y2))))
```

Let's fit separate models to these two curves. This lets the linear and
quadratic effects be estimated separately. But note that the second curve is
not centered on x = 0 as the first curve is.

```{r}
m1 <- lm(I(y1 + noise) ~ poly(x_centered, 2))
m2 <- lm(I(y2 + noise) ~ poly(x_centered, 2))

par(mfrow = c(2, 1))
arm::coefplot(m1, mar = c(2, 10, 2, 2), xlim = c(-700, 0))
arm::coefplot(m2, mar = c(2, 10, 2, 2), xlim = c(-700, 0))
```

Both the linear and quadratic effects are affected but the linear coefficient
is affected more strongly. (The coefficients ending in 2 are the quadratic
coefficients and the ones ending in 1 are the linear coefficients.)

Now let's fit a model that resembles the ones fit in the paper. We are adding
a factor coefficient that can shift the effect up or down when nutrients are
added. We are also adding a interaction with the linear effect. We are
not adding an interaction with the quadratic effect.

```{r}
par(mfrow = c(1, 1))
m <- lm(y ~ poly(x, 2)[,1] + poly(x, 2)[,2] + nutrients +
  nutrients * poly(x, 2)[,1], data = dat)
arm::coefplot(m, mar = c(2, 10, 2, 2))
```

So now we are forcing them to share a quadratic effect. But in this case we
would be able to detect the effect according to an interaction between the
linear effect and the nutrient variable.
